The Rockerverse: Packages and Applications for Containerisation with R

### 2020-03-09 rockerverse.Rmd
fix typos and grammatical errors

  This articles surveys downstream projects that build upon the Rocker Project images and presents the current state of R packages for managing Docker images and controlling containers.

@@

The R community continues to grow.
This can be seen in the number of new packages on CRAN, which is still on growing exponentially \citep{cran:2019}, but also in the numbers of conferences, open educational resources, meetups, unconferences, and companies that are adopting R, as exemplified by the useR! conference series\footnote{\href{https://www.r-project.org/conferences/}{https://www.r-project.org/conferences/}}, the global growth of the R and R-Ladies user groups\footnote{\href{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}, \href{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}}, or the foundation and impact of the R&nbsp;Consortium\footnote{\href{https://www.r-consortium.org/news/announcements}{https://www.r-consortium.org/news/announcements}, \href{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}}.
These trends cement the role of R as the _lingua&nbsp;franca_ of statistics, data visualisation, and computational research.
The last few years, coinciding with the rise of R, have also seen the rise of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) as a general tool for distributing and deploying of server applications---in fact, Docker can be called the _lingua&nbsp;franca_ of describing computing environments and packaging software.
Combining both these topics, the _Rocker&nbsp;Project_ ([https://www.rocker-project.org/](https://www.rocker-project.org/)) provides Docker images with R (see the next section for more details).
The considerable uptake and continued evolution of the Rocker&nbsp;Project has led to numerous projects that extend or build upon Rocker images, ranging from reproducible\footnote{"Reproducible" in the sense of the \emph{Claerbout/Donoho/Peng} terminology \citep{barba_terminologies_2018}.} research to production deployments.
As such, this article presents what we may call the _Rockerverse_ of projects across all development stages: early demonstrations, working prototypes, and mature products.
We also introduce related activities that connect the R language and environment with other containerisation solutions.
Our main contributions is a coherent picture of the current status of using containers in, with, and for R.

@@

Docker, an application and service provided by the eponymous company, has, in just a few short years, risen to prominence for developing, testing, deploying and distributing computer software \citep[cf.][]{datadog_8_2018,munoz_history_2019}.
While related approaches exist, such as LXC\footnote{\href{https://en.wikipedia.org/wiki/LXC}{https://en.wikipedia.org/wiki/LXC}} or Singularity \citep{kurtzer_singularity_2017}, Docker has become synomymous with "containerisation"---the method of taking software artefacts and bundling them in such a way that use becomes standardized and portable across operating systems.
In doing so, Docker had recognised and validated the importance of one very important thread that had been emerging, namely virtualisation.
By allowing (one or possibly) multiple applications or services to run concurrently on one host machine without any fear of interference between them, Docker provides an important scalability opportunity.
Beyond this though, Docker has improved this compartmentalisation by accessing the host system---generally Linux---through a much thinner and smaller shim than a full operating system emulation or virtualisation.
This containerisation, also called operating-system-level virtualisation \citep{wikipedia_contributors_os-level_2020},
makes efficient use of operating system resources \citep{felter_updated_2015} and allows another order of magnitude in terms of scalability of deployment \citep[cf.][]{datadog_8_2018}, because virtualisation may emulate a whole operating system, a container typically runs only one process.
The single process together with sharing the host's kernel results in a reduced footprint and faster start times.
While Docker makes use of Linux kernel features, it has become important enough that some required aspects of running Docker have been added to other operating systems so that those systems can more efficiently support Docker \citep{microsoft_linux_2019}.
The success of Docker has even paved the way for industry collaboration and standardisation \citep{oci_open_2019}.

@@

For statistical computing and analysis centered around R, the **Rocker&nbsp;Project** has provided a variety of Docker containers since it began in 2014 \citep{RJ-2017-065}.
The Rocker&nbsp;Project provides several lines of containers spanning from building blocks with `R-release` or `R-devel`, via containers with [RStudio&nbsp;Server](https://rstudio.com/products/rstudio/) and [Shiny&nbsp;Server](https://rstudio.com/products/shiny/shiny-server/), to domain-specific containers such as [`rocker/geospatial`](https://github.com/rocker-org/geospatial) \citep{rocker_geospatial_2019}.
These containers form _image stacks_, building on top of each other for easier maintainability (i.e., smaller `Dockerfile`s), better composability, and to reduce build time.
Also of note is a series of "versioned" containers which match the R release they contain with the _then-current_ set of packages via the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}.
The Rocker&nbsp;Project's impact and importance was acknowledged by the Chan Zuckerberg Initiative's _Essential Open Source Software for Science_, which provides funding for the projects's sustainable maintenance, community growth, and targeting new hardware platforms including GPUs \citep{chan_zuckerberg_initiative_maintaining_2019}.

Docker is not the only containerisation software.
**Singularity** stems from the domain of high-perfomance computing \citep{kurtzer_singularity_2017} and can also run Docker images.
Rocker images work out of the box if the main process is R, e.g., in `rocker/r-base`, but Singularity does not succeed in running images where there is an init script, e.g., in containers that by default run RStudio&nbsp;Server.
In the latter case, a `Singularity` file, a recipe akin to a `Dockerfile`, needs to be used to make necessary adjustments.
To date, no comparable image stack to the Rocker Project's images exists on [Singularity Hub](https://singularity-hub.org/).
A further tool for running containers is [**podman**](https://github.com/containers/libpod), which can also build `Dockerfile`s and run Docker images.
Proof of concepts exists for using podman to build and run Rocker containers\footnote{See \href{https://github.com/nuest/rodman}{https://github.com/nuest/rodman} and \href{https://github.com/rocker-org/rocker-versioned/issues/187}{https://github.com/rocker-org/rocker-versioned/issues/187}}, but the prevelance of Docker, especially in the broader user community beyond experts or niche systems and the vast amount of blog posts and courses for Docker currently cap specific development efforts for both Singularity and podman in the R community.
This might quickly change if the usability and spread of Singularity or podman increase, or if security features such as rootless/unprivileged containers, which both these tools support out of the box, become more sought after.

@@

**\pkg{harbor}** ([https://github.com/wch/harbor](https://github.com/wch/harbor)) is no longer actively maintained, but it should be honorably mentioned as the first R package for managing Docker images and containers.
It uses the \CRANpkg{sys} package \citep{cran_sys} to run system commands against the Docker CLI, both locally and through an SSH connection, and it has convenience functions, e.g., for listing and removing containers/images and for accessing logs.
The outputs of container executions are converted to appropriate R types.
The Docker CLI's basic functionality, although it evolves quickly and with little concern for avoiding breaking changes, has remained unchanged in core functions, meaning that a core function such as \code{harbor::docker\_run(image = "hello-world")} still works depsite its stopped development.

@@

**\pkg{dockermachine}** ([https://github.com/cboettig/dockermachine](https://github.com/cboettig/dockermachine)) is an R package to provide a convenient interface to [Docker&nbsp;Machine](https://docs.docker.com/machine/overview/) from R.
The CLI tool `docker-machine` allows users to create and manage a virtual host on local computers, local data centers, or at cloud providers.
A local Docker installation can be configured to transparntly forward all commands issued on the local Docker CLI to a selected (remote) virtual host.
Docker&nbsp;Machine was especially crucial for local use in the early days of Docker, when no native support was available for Mac or Windows computers, but it remains relevant for provisioning on remote systems.
The package has not received any updates for two years, but it is functional with a current version of `docker-machine` (`0.16.2`).
It potentially lowers the barriers for R users to run containers on various hosts if they perceive that using the Docker&nbsp;Machine CLI directly as a barrier and it enables scripted workflows with remote processing.

@@

**Bioconductor** ([https://bioconductor.org/](https://bioconductor.org/)) is an open source, open development project for the analysis and comprehension of genomic data \citep{gentleman_bioconductor_2004}.
As of October 30th 2019, the project consists of 1823 R software packages, as well as packages containing annotation or experiment data.
_Bioconductor_ has a semi-annual release cycle, where each release is associated with a particular version of R, and Docker images are provided for current and past versions of _Bioconductor_ for convenience and reproducibility.
All images, which are described on the _Bioconductor_ web site (see [https://bioconductor.org/help/docker/](https://bioconductor.org/help/docker/)), are created with `Dockerfile`s maintained on GitHub and distributed through Docker&nbsp;Hub\footnote{See \href{https://github.com/Bioconductor/bioconductor_docker}{https://github.com/Bioconductor/bioconductor\_docker} and \href{https://hub.docker.com/u/bioconductor}{https://hub.docker.com/u/bioconductor} respectively.}.
_Bioconductor_'s "base" Docker images are built on top of the `rocker/rstudio` image.
_Bioconductor_ installs packages based on the R version in combination with the Bioconductor version and, therefore, uses Bioconductor version tagging `devel` and `RELEASE_X_Y`, e.g., `RELEASE_3_10`.
Past and current combinations of R and _Bioconductor_ will therefore be accessible via a specific image tags.

The _Bioconductor_ `Dockerfile` selects the desired R version from Rocker images, adds required system dependencies, and uses the \CRANpkg{BiocManager} package for installing appropriate versions of _Bioconductor_ packages \citep{cran_biocmanager}.
A strength of this approach is that the responsibility for complex software configuration and customization is shifted from the user to the experienced _Bioconductor_ core team.
However, a recent audit of the _Bioconductor_  image stack `Dockerfile` led to the deprecation of several community-maintained images, because the numerous specific images became too hard to understand, complex to maintain, and cumbersome to customise.
As part of the simplification, a recent innovation is the `bioconductor_docker:devel` image, which emulates the _Bioconductor_ environment for nightly builds as closely as possible.
This image contains the environment variables and the system dependencies needed to install and check almost all _Bioconductor_ software packages (1813 out of 1823).
It saves users and package developers from creating this environment themselves.
Furthermore, the image is configured so that \code{.libPaths()} has \samp{/usr/local/lib/R/host-site-library} as the first location.
Users mounting a location on the host file system to this location can persistently manage installed packages across Docker containers or image updates.
Many R users pursue flexible work flows tailored to particular analysis needs rather than standardized work flows.
The new `bioconductor_docker` image is well suited for this preference, while `bioconductor_docker:devel` provides developers with a test environment close to _Bioconductor_'s build system.

\label{datascience}
**Data science** is a widely discussed topic in all academic disciplines \citep[e.g.,][]{donoho_50_2017}.
These discussions have shed a light on the tools and craftspersonship behind the analysis of data with computational methods.
The practice of data science often involves combining tools and software stacks and requires a cross-cutting skillset.
This complexity and an inherent concern for openness and reproducibility in the data science community has led to Docker being used widely.
The remainder of this section presents example Docker images and image stacks featuring R intended for data science.

@@

The R images are based on Gigantum's `python3-minimal` image, originally to keep the existing front-end configuration, but also to provide a consistent Python-to-R interoperability.

@@

Both approaches enable the `apt` package manager \citep{wikipedia_contributors_apt_2020} as an installation method, and this is exposed via the GUI-based environment management,\footnote{See \href{https://docs.gigantum.com/docs/environment-management}{https://docs.gigantum.com/docs/environment-management}}.

@@

**\pkg{containerit}** ([https://github.com/o2r-project/containerit/](https://github.com/o2r-project/containerit/)) attempts to take this one step further and includes these tools to automatically create a `Dockerfile` that can execute a given workflow \citep{nust_containerit_2019}.
\pkg{containerit} accepts an R object of classes \code{"sessionInfo"} or \code{"session\_info"} as input and provides helper functions to derive these from workflows, e.g., an R script or R Markdown document, by analysing the session state at the end of the workflow.
It relies on the \pkg{sysreqs} ([https://github.com/r-hub/sysreqs/](https://github.com/r-hub/sysreqs/)) package and it's mapping of package system dependencies to platform specific installation package names\footnote{See \href{https://sysreqs.r-hub.io/}{https://sysreqs.r-hub.io/}.}.
\pkg{containerit} uses \pkg{stevedore} to streamline the user interaction and improve the created `Dockerfile`s, e.g., by running a container for the desired base image to extract the already available R packages.

@@

**\CRANpkg{liftr}** \citep{liftr2019} aims to solve the problem of persistent reproducible reporting in statistical computing based on the R Markdown format \citep{xie2018}.
The irreproducibility of authoring environments can become an issue for collaborative documents and large-scale platforms for processing documents.
\pkg{liftr} makes the dynamic R Markdown document the main and sole workflow control file and the only file that needs to be shared between collaborators for consistent environments, e.g. demonstrated in the DockFlow project (https://dockflow.org).
It introduces new fields to the document header, allowing users to manually declare the versioned dependencies required for rendering the document.
The package then generates a `Dockerfile` from this metadata and provides a utitility function to render the document inside a Docker container, i.e., \code{render\_docker("foo.Rmd")}.
An RStudio addin even allows compilation of documents with the single push of a button.

System dependencies are the domain of Docker, but for a full description of the computing environment one must also manage the R version and the the R packages.
R versions are available via the versioned Rocker image stack \citep{RJ-2017-065}.
[r-online](https://github.com/ColinFay/ronline) leverages these images and provides an app for helping users to detect breaking changes between different R versions and for historic exploration of R.
With a standalone NodeJS app or [r-online](https://srv.colinfay.me/r-online), the user can compare a piece of code run in two separate versions of R.
Internally, r-online opens one or two Docker instances with the given version of R based on Rocker images, executes a given piece of code, and returns the result to the user.
Regarding R package management, this can be achieved with MRAN, or with packages such as \CRANpkg{checkpoint} \citep{cran_checkpoint} and \CRANpkg{renv} \citep{renv2019}, which can naturally be applied within images and containers.
For example, \pkg{renv} helps users to manage the state of the R library in a reproducible way, further providing isolation and portability.
While \pkg{renv} does not cover system dependencies, the \pkg{renv}-based environment can be transferred into a container either by restoring the environment based on the main configuration file `renv.lock` or by storing the \pkg{renv}-cache on the host and not in the container \citep{ushey_using_2019}.
With both the system dependencies and R packages consciously managed in a Docker image, users can start using containers as the _only_ environment for their workflows, which allows them to work indepenently of physical computers\footnote{Allowing them to be digital "nomads", cf. J.~Bryan's \href{https://github.com/jennybc/docker-why}{https://github.com/jennybc/docker-why}.} and to assert a specific degree of confidence in the stability of a developed software \citep[cf. \texttt{README.Rmd} in][]{marwick_research_2017}.

@@

Third, special-purpose images exist for identifying problems beyond the mere R code, such as **debugging R memory problems**.
These images significantly reduce the barriers to following complex steps for fixing memory allocation bugs \citep[cf. Section~4.3 in][]{core_writing_1999}.
These problems are hard to debug and critical, because when they do occur they lead to fatal crashing processes.
[`rocker/r-devel-san`](https://github.com/rocker-org/r-devel-san) and [`rocker/r-devel-ubsan-clang`](https://github.com/rocker-org/r-devel-san-clang) are Docker images that have a particularly configured version of R to trace such problems with gcc and clang compilers, respectively \citep[cf.~\CRANpkg{sanitizers} for examples,][]{eddelbuettel_sanitizers_2014}.
[`wch/r-debug`](https://github.com/wch/r-debug) is a purpose-built Docker image with _multiple_ instrumented builds of R, each with a different diagnostic utility activated.

Fourth, containers are useful for **testing** R code during development.
To submit a package to CRAN, an R package must work with the development version of R, which must be compiled locally; this can be a challenge for some users.
The **R-hub** project provides _"a collection of services to help R package development"_, with the package builder as the most prominent one \citep{r-hub_docs_2019}.
R-hub makes it easy to ensure that no errors occur, but fixing errors still often warrants a local setup, e.g., using the image `rocker/r-devel`, as is testing packages with native code, which can make the process more complex \citep[cf.][]{eckert_building_2018}.
The R-hub Docker images can also be used to debug problems locally using various combinations of Linux platforms, R versions, and compilers\footnote{See \href{https://r-hub.github.io/rhub/articles/local-debugging.html}{https://r-hub.github.io/rhub/articles/local-debugging.html} and \href{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}}.
The images go beyond the configurations, or _flavours_, used by CRAN for checking packages\footnote{\href{https://cran.r-project.org/web/checks/check_flavors.html}{https://cran.r-project.org/web/checks/check\_flavors.html}}, e.g., with CentOS-based images, but they lack a container for checking on Windows or OS X.
The images greatly support package developers to provide support on operating systems with which they are not familiar.
The package \pkg{dockertest} ([https://github.com/traitecoevo/dockertest/](https://github.com/traitecoevo/dockertest/)) is a proof of concept for automatically generating `Dockerfile`s and building images specifically to run tests\footnote{\pkg{dockertest} is not actively maintained, but mentioned still because of its interesting approach.}.
These images are accompanied by a special launch script so the tested source code is not stored in the image; instead, the currently checked in version from a local Git repository is cloned into the container at runtime.
This approach clearly separates test environment, test code, and current working copy of the code.
\label{rselenium} Another use case where a container can help to standardise tests across operating systems is detailed the vignettes of the package \CRANpkg{RSelenium} \citep{rselenium_2019}.
The package recommends Docker for running the [Selenium](https://selenium.dev/) Server application needed to execute test suites on browser-based user interfaces and webpages, but it requires users to manually manage the containers.

>>>>

  This article surveys downstream projects that build upon the Rocker Project images and presents the current state of R packages for managing Docker images and controlling containers.

@@

The R community continues to grow.
This can be seen in the number of new packages on CRAN, which is still on growing exponentially \citep{cran:2019}, but also in the numbers of conferences, open educational resources, meetups, unconferences, and companies that are adopting R, as exemplified by the useR! conference series\footnote{\href{https://www.r-project.org/conferences/}{https://www.r-project.org/conferences/}}, the global growth of the R and R-Ladies user groups\footnote{\href{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}, \href{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}}, or the foundation and impact of the R&nbsp;Consortium\footnote{\href{https://www.r-consortium.org/news/announcements}{https://www.r-consortium.org/news/announcements}, \href{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}}.
These trends cement the role of R as the _lingua&nbsp;franca_ of statistics, data visualisation, and computational research.
The last few years, coinciding with the rise of R, have also seen the rise of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) as a general tool for distributing and deploying of server applications---in fact, Docker can be called the _lingua&nbsp;franca_ of describing computing environments and packaging software.
Combining both these topics, the _Rocker&nbsp;Project_ ([https://www.rocker-project.org/](https://www.rocker-project.org/)) provides Docker images with R (see the next section for more details).
The considerable uptake and continued evolution of the Rocker&nbsp;Project has led to numerous projects that extend or build upon Rocker images, ranging from reproducible\footnote{"Reproducible" in the sense of the \emph{Claerbout/Donoho/Peng} terminology \citep{barba_terminologies_2018}.} research to production deployments.
As such, this article presents what we may call the _Rockerverse_ of projects across all development stages: early demonstrations, working prototypes, and mature products.
We also introduce related activities that connect the R language and environment with other containerisation solutions.
Our main contribution is a coherent picture of the current status of using containers in, with, and for R.

@@

Docker, an application and service provided by the eponymous company, has, in just a few short years, risen to prominence for developing, testing, deploying and distributing computer software \citep[cf.][]{datadog_8_2018,munoz_history_2019}.
While related approaches exist, such as LXC\footnote{\href{https://en.wikipedia.org/wiki/LXC}{https://en.wikipedia.org/wiki/LXC}} or Singularity \citep{kurtzer_singularity_2017}, Docker has become synonymous with "containerisation"---the method of taking software artefacts and bundling them in such a way that use becomes standardized and portable across operating systems.
In doing so, Docker had recognised and validated the importance of one very important thread that had been emerging, namely virtualisation.
By allowing (one or possibly) multiple applications or services to run concurrently on one host machine without any fear of interference between them, Docker provides an important scalability opportunity.
Beyond this though, Docker has improved this compartmentalisation by accessing the host system---generally Linux---through a much thinner and smaller shim than a full operating system emulation or virtualisation.
This containerisation, also called operating-system-level virtualisation \citep{wikipedia_contributors_os-level_2020},
makes efficient use of operating system resources \citep{felter_updated_2015} and allows another order of magnitude in terms of scalability of deployment \citep[cf.][]{datadog_8_2018}, because virtualisation may emulate a whole operating system, a container typically runs only one process.
The single process together with sharing the host's kernel results in a reduced footprint and faster start times.
While Docker makes use of Linux kernel features, it has become important enough that some required aspects of running Docker have been added to other operating systems so that those systems can more efficiently support Docker \citep{microsoft_linux_2019}.
The success of Docker has even paved the way for industry collaboration and standardisation \citep{oci_open_2019}.

@@

For statistical computing and analysis centred around R, the **Rocker&nbsp;Project** has provided a variety of Docker containers since it began in 2014 \citep{RJ-2017-065}.
The Rocker&nbsp;Project provides several lines of containers spanning from building blocks with `R-release` or `R-devel`, via containers with [RStudio&nbsp;Server](https://rstudio.com/products/rstudio/) and [Shiny&nbsp;Server](https://rstudio.com/products/shiny/shiny-server/), to domain-specific containers such as [`rocker/geospatial`](https://github.com/rocker-org/geospatial) \citep{rocker_geospatial_2019}.
These containers form _image stacks_, building on top of each other for easier maintainability (i.e., smaller `Dockerfile`s), better composability, and to reduce build time.
Also of note is a series of "versioned" containers which match the R release they contain with the _then-current_ set of packages via the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}.
The Rocker&nbsp;Project's impact and importance was acknowledged by the Chan Zuckerberg Initiative's _Essential Open Source Software for Science_, which provides funding for the project's sustainable maintenance, community growth, and targeting new hardware platforms including GPUs \citep{chan_zuckerberg_initiative_maintaining_2019}.

Docker is not the only containerisation software.
**Singularity** stems from the domain of high-performance computing \citep{kurtzer_singularity_2017} and can also run Docker images.
Rocker images work out of the box if the main process is R, e.g., in `rocker/r-base`, but Singularity does not succeed in running images where there is an init script, e.g., in containers that by default run RStudio&nbsp;Server.
In the latter case, a `Singularity` file, a recipe akin to a `Dockerfile`, needs to be used to make necessary adjustments.
To date, no comparable image stack to the Rocker Project's images exists on [Singularity Hub](https://singularity-hub.org/).
A further tool for running containers is [**podman**](https://github.com/containers/libpod), which can also build `Dockerfile`s and run Docker images.
Proof of concepts exists for using podman to build and run Rocker containers\footnote{See \href{https://github.com/nuest/rodman}{https://github.com/nuest/rodman} and \href{https://github.com/rocker-org/rocker-versioned/issues/187}{https://github.com/rocker-org/rocker-versioned/issues/187}}, but the prevalence of Docker, especially in the broader user community beyond experts or niche systems and the vast amount of blog posts and courses for Docker currently cap specific development efforts for both Singularity and podman in the R community.
This might quickly change if the usability and spread of Singularity or podman increase, or if security features such as rootless/unprivileged containers, which both these tools support out of the box, become more sought after.

@@

**\pkg{harbor}** ([https://github.com/wch/harbor](https://github.com/wch/harbor)) is no longer actively maintained, but it should be honourably mentioned as the first R package for managing Docker images and containers.
It uses the \CRANpkg{sys} package \citep{cran_sys} to run system commands against the Docker CLI, both locally and through an SSH connection, and it has convenience functions, e.g., for listing and removing containers/images and for accessing logs.
The outputs of container executions are converted to appropriate R types.
The Docker CLI's basic functionality, although it evolves quickly and with little concern for avoiding breaking changes, has remained unchanged in core functions, meaning that a core function such as \code{harbor::docker\_run(image = "hello-world")} still works despite its stopped development.

@@

**\pkg{dockermachine}** ([https://github.com/cboettig/dockermachine](https://github.com/cboettig/dockermachine)) is an R package to provide a convenient interface to [Docker&nbsp;Machine](https://docs.docker.com/machine/overview/) from R.
The CLI tool `docker-machine` allows users to create and manage a virtual host on local computers, local data centres, or at cloud providers.
A local Docker installation can be configured to transparently forward all commands issued on the local Docker CLI to a selected (remote) virtual host.
Docker&nbsp;Machine was especially crucial for local use in the early days of Docker, when no native support was available for Mac or Windows computers, but it remains relevant for provisioning on remote systems.
The package has not received any updates for two years, but it is functional with a current version of `docker-machine` (`0.16.2`).
It potentially lowers the barriers for R users to run containers on various hosts if they perceive that using the Docker&nbsp;Machine CLI directly as a barrier and it enables scripted workflows with remote processing.

@@

**Bioconductor** ([https://bioconductor.org/](https://bioconductor.org/)) is an open-source, open development project for the analysis and comprehension of genomic data \citep{gentleman_bioconductor_2004}.
As of October 30th 2019, the project consists of 1823 R software packages, as well as packages containing annotation or experiment data.
_Bioconductor_ has a semi-annual release cycle, where each release is associated with a particular version of R, and Docker images are provided for current and past versions of _Bioconductor_ for convenience and reproducibility.
All images, which are described on the _Bioconductor_ web site (see [https://bioconductor.org/help/docker/](https://bioconductor.org/help/docker/)), are created with `Dockerfile`s maintained on GitHub and distributed through Docker&nbsp;Hub\footnote{See \href{https://github.com/Bioconductor/bioconductor_docker}{https://github.com/Bioconductor/bioconductor\_docker} and \href{https://hub.docker.com/u/bioconductor}{https://hub.docker.com/u/bioconductor} respectively.}.
_Bioconductor_'s "base" Docker images are built on top of the `rocker/rstudio` image.
_Bioconductor_ installs packages based on the R version in combination with the Bioconductor version and, therefore, uses Bioconductor version tagging `devel` and `RELEASE_X_Y`, e.g., `RELEASE_3_10`.
Past and current combinations of R and _Bioconductor_ will therefore be accessible via specific image tags.

The _Bioconductor_ `Dockerfile` selects the desired R version from Rocker images, adds required system dependencies, and uses the \CRANpkg{BiocManager} package for installing appropriate versions of _Bioconductor_ packages \citep{cran_biocmanager}.
A strength of this approach is that the responsibility for complex software configuration and customization is shifted from the user to the experienced _Bioconductor_ core team.
However, a recent audit of the _Bioconductor_  image stack `Dockerfile` led to the deprecation of several community-maintained images, because the numerous specific images became too hard to understand, complex to maintain, and cumbersome to customise.
As part of the simplification, a recent innovation is the `bioconductor_docker:devel` image, which emulates the _Bioconductor_ environment for nightly builds as closely as possible.
This image contains the environment variables and the system dependencies needed to install and check almost all _Bioconductor_ software packages (1813 out of 1823).
It saves users and package developers from creating this environment themselves.
Furthermore, the image is configured so that \code{.libPaths()} has \samp{/usr/local/lib/R/host-site-library} as the first location.
Users mounting a location on the host file system to this location can persistently manage installed packages across Docker containers or image updates.
Many R users pursue flexible workflows tailored to particular analysis needs rather than standardized workflows.
The new `bioconductor_docker` image is well suited for this preference, while `bioconductor_docker:devel` provides developers with a test environment close to _Bioconductor_'s build system.

\label{datascience}
**Data science** is a widely discussed topic in all academic disciplines \citep[e.g.,][]{donoho_50_2017}.
These discussions have shed light on the tools and craftspersonship behind the analysis of data with computational methods.
The practice of data science often involves combining tools and software stacks and requires a cross-cutting skillset.
This complexity and an inherent concern for openness and reproducibility in the data science community has led to Docker being used widely.
The remainder of this section presents example Docker images and image stacks featuring R intended for data science.

@@

The R images are based on Gigantum's `python3-minimal` image, originally to keep the existing front-end configuration, but also to provide consistent Python-to-R interoperability.

@@

Both approaches enable the `apt` package manager \citep{wikipedia_contributors_apt_2020} as an installation method, and this is exposed via the GUI-based environment management\footnote{See \href{https://docs.gigantum.com/docs/environment-management}{https://docs.gigantum.com/docs/environment-management}}.

@@

**\pkg{containerit}** ([https://github.com/o2r-project/containerit/](https://github.com/o2r-project/containerit/)) attempts to take this one step further and includes these tools to automatically create a `Dockerfile` that can execute a given workflow \citep{nust_containerit_2019}.
\pkg{containerit} accepts an R object of classes \code{"sessionInfo"} or \code{"session\_info"} as input and provides helper functions to derive these from workflows, e.g., an R script or R Markdown document, by analysing the session state at the end of the workflow.
It relies on the \pkg{sysreqs} ([https://github.com/r-hub/sysreqs/](https://github.com/r-hub/sysreqs/)) package and it's mapping of package system dependencies to platform-specific installation package names\footnote{See \href{https://sysreqs.r-hub.io/}{https://sysreqs.r-hub.io/}.}.
\pkg{containerit} uses \pkg{stevedore} to streamline the user interaction and improve the created `Dockerfile`s, e.g., by running a container for the desired base image to extract the already available R packages.

@@

**\CRANpkg{liftr}** \citep{liftr2019} aims to solve the problem of persistent reproducible reporting in statistical computing based on the R Markdown format \citep{xie2018}.
The irreproducibility of authoring environments can become an issue for collaborative documents and large-scale platforms for processing documents.
\pkg{liftr} makes the dynamic R Markdown document the main and sole workflow control file and the only file that needs to be shared between collaborators for consistent environments, e.g. demonstrated in the DockFlow project (https://dockflow.org).
It introduces new fields to the document header, allowing users to manually declare the versioned dependencies required for rendering the document.
The package then generates a `Dockerfile` from this metadata and provides a utility function to render the document inside a Docker container, i.e., \code{render\_docker("foo.Rmd")}.
An RStudio addin even allows compilation of documents with the single push of a button.

System dependencies are the domain of Docker, but for a full description of the computing environment, one must also manage the R version and the R packages.
R versions are available via the versioned Rocker image stack \citep{RJ-2017-065}.
[r-online](https://github.com/ColinFay/ronline) leverages these images and provides an app for helping users to detect breaking changes between different R versions and for historic exploration of R.
With a standalone NodeJS app or [r-online](https://srv.colinfay.me/r-online), the user can compare a piece of code run in two separate versions of R.
Internally, r-online opens one or two Docker instances with the given version of R based on Rocker images, executes a given piece of code, and returns the result to the user.
Regarding R package management, this can be achieved with MRAN, or with packages such as \CRANpkg{checkpoint} \citep{cran_checkpoint} and \CRANpkg{renv} \citep{renv2019}, which can naturally be applied within images and containers.
For example, \pkg{renv} helps users to manage the state of the R library in a reproducible way, further providing isolation and portability.
While \pkg{renv} does not cover system dependencies, the \pkg{renv}-based environment can be transferred into a container either by restoring the environment based on the main configuration file `renv.lock` or by storing the \pkg{renv}-cache on the host and not in the container \citep{ushey_using_2019}.
With both the system dependencies and R packages consciously managed in a Docker image, users can start using containers as the _only_ environment for their workflows, which allows them to work independently of physical computers\footnote{Allowing them to be digital "nomads", cf. J.~Bryan's \href{https://github.com/jennybc/docker-why}{https://github.com/jennybc/docker-why}.} and to assert a specific degree of confidence in the stability of a developed software \citep[cf. \texttt{README.Rmd} in][]{marwick_research_2017}.

@@

Third, special-purpose images exist for identifying problems beyond the mere R code, such as **debugging R memory problems**.
These images significantly reduce the barriers to following complex steps for fixing memory allocation bugs \citep[cf. Section~4.3 in][]{core_writing_1999}.
These problems are hard to debug and critical, because when they do occur they lead to fatal crashes.
[`rocker/r-devel-san`](https://github.com/rocker-org/r-devel-san) and [`rocker/r-devel-ubsan-clang`](https://github.com/rocker-org/r-devel-san-clang) are Docker images that have a particularly configured version of R to trace such problems with gcc and clang compilers, respectively \citep[cf.~\CRANpkg{sanitizers} for examples,][]{eddelbuettel_sanitizers_2014}.
[`wch/r-debug`](https://github.com/wch/r-debug) is a purpose-built Docker image with _multiple_ instrumented builds of R, each with a different diagnostic utility activated.

Fourth, containers are useful for **testing** R code during development.
To submit a package to CRAN, an R package must work with the development version of R, which must be compiled locally; this can be a challenge for some users.
The **R-hub** project provides _"a collection of services to help R package development"_, with the package builder as the most prominent one \citep{r-hub_docs_2019}.
R-hub makes it easy to ensure that no errors occur, but fixing errors still often warrants a local setup, e.g., using the image `rocker/r-devel`, as is testing packages with native code, which can make the process more complex \citep[cf.][]{eckert_building_2018}.
The R-hub Docker images can also be used to debug problems locally using various combinations of Linux platforms, R versions, and compilers\footnote{See \href{https://r-hub.github.io/rhub/articles/local-debugging.html}{https://r-hub.github.io/rhub/articles/local-debugging.html} and \href{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}}.
The images go beyond the configurations, or _flavours_, used by CRAN for checking packages\footnote{\href{https://cran.r-project.org/web/checks/check_flavors.html}{https://cran.r-project.org/web/checks/check\_flavors.html}}, e.g., with CentOS-based images, but they lack a container for checking on Windows or OS X.
The images greatly support package developers to provide support on operating systems with which they are not familiar.
The package \pkg{dockertest} ([https://github.com/traitecoevo/dockertest/](https://github.com/traitecoevo/dockertest/)) is a proof of concept for automatically generating `Dockerfile`s and building images specifically to run tests\footnote{\pkg{dockertest} is not actively maintained, but mentioned still because of its interesting approach.}.
These images are accompanied by a special launch script so the tested source code is not stored in the image; instead, the currently checked in version from a local Git repository is cloned into the container at runtime.
This approach separates the test environment, test code, and current working copy of the code.
\label{rselenium} Another use case where a container can help to standardise tests across operating systems is detailed the vignettes of the package \CRANpkg{RSelenium} \citep{rselenium_2019}.
The package recommends Docker for running the [Selenium](https://selenium.dev/) Server application needed to execute test suites on browser-based user interfaces and webpages, but it requires users to manually manage the containers.
